---
title: "00_rbf testing"
author: "Brenna Kelly"
date: "2025-11-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Objectives:  
- demonstrate the use of spatial information, why we might include it; how spatial information is encoded  
- what is the task? prediction or understanding  
- what different types of spatial information add   
- how do RBFs work  
- can improve predictions but also improve understanding (account for some portion of spatial autocorrelation)  

```{r}

library(sp)
library(sf)
library(tmap)
library(mlr3)
library(ranger)
library(mlr3learners)

data("meuse")

```



```{r}

meuse <- na.omit(meuse)

meuse$id <- row.names(meuse)

plot(cbind(st_as_sf(meuse, coords = c("x", "y"), crs = 28992),
           meuse$x, meuse$y), max.plot = 15)


```

### Testing  
We'll try a few different scenarios: no explicitly spatial information, some spatial information, only spatial information

```{r}

all_data <- meuse

no_spatial_subset <- subset(meuse, select = -c(x, y, dist, dist.m))

# some spatial
no_coords_subset <- subset(meuse, select = -c(x, y))
some_spatial_subset <- subset(meuse, select = -c(dist, dist.m))

# only coords
coords_subset <- subset(meuse, select = c(x, y, id, zinc))
# coords and distance to river
coords_dist_subset <- subset(meuse, select = c(x, y, id, dist, dist.m, zinc))

# distance between points
meuse_dist <- as.matrix(dist(cbind(meuse$x, meuse$y)))
meuse_dist_inv <- 1 / meuse_dist
diag(meuse_dist_inv) <- 0

meuse_idw <- cbind(meuse, meuse_dist_inv)
excl_names <- names(subset(meuse, select = -c(zinc, id)))
idw_subset <- meuse_idw[, !names(meuse_idw) %in% excl_names]

# distance plus other non-spatial
idw_plus_subset <- subset(meuse_idw, select = -c(x, y, dist, dist.m))
# idw_plus_subset <- subset(meuse_idw, select = -c(dist, dist.m))

data_subsets <- list(all_data, no_spatial_subset, no_coords_subset,
                     some_spatial_subset, coords_subset,
                     coords_dist_subset, idw_subset,
                     meuse_idw)
names(data_subsets) <- c("all_data",
                         "no_spatial_subset", "no_coords_subset",
                         "some_spatial_subset", "coords_subset",
                         "coords_dist_subset", "idw_subset",
                         "idw_plus_subset")

```


### Function to return RF results  
Takes a dataframe which contains `zinc` and the desired covariates.

```{r}

rf_fx <- function(data) {
  
  # task
  task <- TaskRegr$new("id", 
                       subset(data, select = -c(id)), 
                       target = "zinc")
  # learner
  learn_rf <- lrn("regr.ranger", importance = "permutation")
  
  # train
  learn_rf$train(task)
  
  # predict
  predictions <- learn_rf$predict(task)
  
  print(
    predictions$score(msr("regr.rmse"))
  )
  
  print(learn_rf$importance())
  # print(learn_rf$selected_features())
  
  # meuse_res <- cbind(meuse, y_hat = predictions$response)
  # meuse_res$residual <- meuse_res$zinc - meuse_res$y_hat
  # 
  # print(moranfast(meuse_res$residual, meuse_res$x, meuse_res$y))
  
}

for(i in 1:length(data_subsets)) {
  
  print(names(data_subsets[i]))
  rf_fx(data_subsets[[i]])
  
}

```




### An aspatial random forest

```{r}

# train_dat <- trainControl(method = "cv", number = 5)

# task
task <- TaskRegr$new("id", 
                     subset(idw_subset, select = -c(id)), 
                     target = "zinc")

# learner
learn_rf <- lrn("regr.ranger", importance = "permutation")

# learn_rf$param_set$values$maxdepth

# train
learn_rf$train(task)

# predict
predictions <- learn_rf$predict(task)

# performance
predictions$score(msr("regr.rmse"))
sd(meuse$zinc)

# plot predictions

meuse_res <- cbind(meuse, y_hat = predictions$response)
meuse_res$residual <- meuse_res$zinc - meuse_res$y_hat

# make sf
meuse_sf <- st_as_sf(meuse, coords = c("x", "y"), crs = 28992)

tmap_mode(mode = "view")

pred_zinc <- tm_shape(meuse_res_sf) +
  tm_dots(fill = "y_hat", fill.scale = tm_scale_continuous(values = "viridis",
                                                           limits = c(110, 1850)),
          size = 0.5) +
  tm_layout(frame = FALSE,
            legend.frame = FALSE)

true_zinc <- tm_shape(meuse_res_sf) +
  tm_dots(fill = "zinc", fill.scale = tm_scale_continuous(values = "viridis",
                                                          limits = c(110, 1850)),
          size = 0.5) +
  tm_layout(frame = FALSE,
            legend.frame = FALSE)

tmap_arrange(true_zinc, pred_zinc)

# plot residuals
tm_shape(meuse_res_sf) +
  tm_dots(fill = "residual", fill.scale = tm_scale_continuous(),
          size = 0.5) +
  tm_layout(frame = FALSE,
            legend.frame = FALSE)

# is there spatial autocorrelation in the residuals?

# meuse_dist <- as.matrix(dist(cbind(meuse_res$x, meuse_res$y)))
# meuse_dist_inv <- 1 / meuse_dist
# diag(meuse_dist_inv) <- 0

library(moranfast)
moranfast(meuse_res$residual, meuse_res$x, meuse_res$y)

# what features did it use?
learn_rf$selected_features()

```


Let's say we like this prediction. What if we want to extrapolate?

```{r}

# define the centroids of a grid to extrapolate to
# plot(st_make_grid(st_bbox(meuse_sf), cellsize = 100, square = FALSE))

grid <- st_make_grid(st_bbox(meuse_sf), cellsize = 100, square = FALSE)
grid_centers <- st_centroid(grid)
grid_centers <- st_as_sf(grid_centers)
names(grid_centers) <- "geometry"

grid <- st_as_sf(grid) # keep, make sf

grid_dist <- cbind(grid_centers, 
                   st_distance(grid_centers$geometry, 
                               meuse_sf$geometry))
names(grid_dist) <- gsub("X", "", names(grid_dist))
grid_dist <- st_drop_geometry(grid_dist)

grid_dist[] <- lapply(grid_dist, as.numeric)
grid_dist[] <- 1 / grid_dist[]

# grid_dist[374, ] # highest around 104; ~0.109
# subset(idw_subset, select = -c(id, zinc))[104, ]

# names(grid_dist) <- paste0("b_", names(grid_dist))

grid_predictions = learn_rf$predict_newdata(grid_dist)


grid$y_hat = grid_predictions$response
grid$id = as.numeric(row.names(grid))

tm_shape(grid) +
  tm_polygons(fill = "y_hat", fill.scale = 
                tm_scale_continuous(values = "viridis"),
              lwd = 0) +
  tm_shape(meuse_sf) +
  tm_dots(fill = "white")

## IDW gives roundness; more flexible functions may be preferred; some bulls-eye behavior

```

### RBFs

```{r}

library(FRK)

?auto_basis

meuse_basis <- auto_basis(data = as_Spatial(meuse_sf),
                          nres = 2, prune = 1,
                          regular = 0, type = "bisquare")

show_basis(meuse_basis) +
  coord_fixed() +
  geom_sf(data = meuse_sf, fill = NA, color = "red")

# evaluate the basis functions at each meuse point
basis_res <- eval_basis(meuse_basis, as_Spatial(meuse_sf))
colnames(basis_res) <- paste0("b_", 1:ncol(basis_res))

# add new features to meuse data
meuse_dat_basis <- cbind(meuse, as.matrix(basis_res))

```

Now test this with the RF

```{r}

rf_fx(meuse_dat_basis) # quick test with all features

# now with just basis functions
excl_names <- names(subset(meuse, select = -c(zinc, id)))
rbf_subset <- meuse_dat_basis[, !names(meuse_dat_basis) %in% excl_names]

task <- TaskRegr$new("id", 
                     subset(rbf_subset, select = -c(id)), 
                     target = "zinc")
# learner
learn_rf <- lrn("regr.ranger", importance = "permutation")

# train
learn_rf$train(task)

# evaluate
predictions$score(msr("regr.rmse"))

# predict
predictions <- learn_rf$predict(task)

predictions$score(msr("regr.rmse"))

learn_rf$importance()

meuse_res <- cbind(meuse, y_hat = predictions$response)
meuse_res$residual <- meuse_res$zinc - meuse_res$y_hat

moranfast(meuse_res$residual, meuse_res$x, meuse_res$y)

```

Now predict at other locations

```{r}

## if using BAUs
# Grid1D_df <- auto_BAUs(manifold = plane(),
#                        cellsize = 200,
#                        type = "hex",
#                        data = as_Spatial(meuse_sf),
#                        nonconvex_hull = 0)
# spplot(Grid1D_df)

grid <- st_make_grid(st_bbox(meuse_sf), cellsize = 100, square = FALSE)
grid_centers <- st_centroid(grid)
# grid_centers <- st_as_sf(grid_centers)
names(grid_centers) <- "geometry"

grid_centers <- as.data.frame(grid_centers)
grid_centers$id <- 1:length(grid_centers)

x_coord_g = st_coordinates(grid_centers)[, "X"]
y_coord_g = st_coordinates(grid_centers)[, "Y"]

grid_pts <- st_as_sf(data.frame("x" = x_coord_g, "y" = y_coord_g),
                     coords = c("x", "y"), crs = 28992)
grid_pts$id <- 1:nrow(grid_pts)


grid_centers <- st_as_sf(grid_centers, geometry = geometry, crs = 28992)
# grid <- st_as_sf(grid)

# evaluate the basis functions at each new point

# test_basis <- auto_basis(data = as_Spatial(grid_pts),
#                           nres = 2, prune = 1,
#                           regular = 0, type = "bisquare")

# test_res <- eval_basis(meuse_basis, as_Spatial(meuse_sf))
grid_basis_res <- eval_basis(meuse_basis, as_Spatial(grid_pts))
colnames(grid_basis_res) <- paste0("b_", 1:ncol(grid_basis_res))

grid_basis_res <- data.frame(as.matrix(grid_basis_res))

# predict for new locations
grid_predictions = learn_rf$predict_newdata(grid_basis_res)

# plot
grid_res <- cbind(st_as_sf(grid), y_hat = grid_predictions$response)

tm_shape(grid_res) +
  tm_polygons(fill = "y_hat", fill.scale = 
                tm_scale_continuous(values = "viridis"),
              lwd = 0) +
  tm_shape(meuse_sf) +
  tm_dots(fill = "white")



```

Do RBFs (in addition to other variables) give us a good model?

```{r}

# now with all variables and basis functions
task <- TaskRegr$new("id", 
                     subset(meuse_dat_basis, select = -c(id)), 
                     target = "zinc")
# learner
learn_rf <- lrn("regr.ranger", importance = "permutation")

# train
learn_rf$train(task)

# predict
predictions <- learn_rf$predict(task)

# evaluate
predictions$score(msr("regr.rmse"))

learn_rf$importance()

# plot final prediction surface
meuse_rbf_res <- cbind(meuse_sf, y_hat = predictions$response)

tm_shape(meuse_rbf_res) +
  tm_dots(fill = "y_hat", fill.scale = 
                tm_scale_continuous(values = "viridis"),
              size = 1)

# is there anymore autocorrelation?
meuse_res <- cbind(meuse, y_hat = predictions$response)
meuse_res$residual <- meuse_res$zinc - meuse_res$y_hat

moranfast(meuse_res$residual, meuse_res$x, meuse_res$y)



```

Is that better than the IDW model with other variables?

```{r}

rf_fx(idw_plus_subset) # RMSE = 63.6 vs. 50 with basis functions

```

Note that the RBF model changes feature importance — cadmium is relatively less important than it was in the original model, and it's also less important than copper.

```{r}
# library(LatticeKrig)
# 
# grid <- st_make_grid(st_bbox(meuse_sf), cellsize = 1000, square = FALSE)
# grid <- st_as_sf(grid)
# 
# # prep the grid
# x_coord_g = st_coordinates(grid)[, "X"] / 100000
# y_coord_g = st_coordinates(grid)[, "Y"] / 100000
# grid_matrix = matrix(c(x_coord_g, y_coord_g), 
#                      nrow = length(y_coord_g), 
#                      ncol = 2, byrow = TRUE)
# 
# # where to evaluate the RBFs (using meuse)
# x_coord_m = st_coordinates(meuse_sf)[, "X"] / 100000
# y_coord_m = st_coordinates(meuse_sf)[, "Y"] / 100000
# meuse_matrix = matrix(c(x_coord_m, y_coord_m), 
#                       nrow = length(y_coord_m), 
#                       ncol = 2, byrow = TRUE)
# 
# distances = rdist(grid_matrix, meuse_matrix)
# wendland_m <- Radial.basis(gl, #matrix(runif(20), ncol = 2), #matrix(x_coord_m, y_coord_m),
#                            meuse_matrix,#matrix(runif(10), ncol = 2), #centers = matrix(x_coord_g, y_coord_g),
#                            d = distances,
#                            # distance.type = "Euclidean",
#                            basis.delta = 0.5, # scale parameter
#                            max.points = 10000,
#                            BasisFunction = "WendlandFunction")
# wendland_m

```



```{r}
# dt<- seq( 0,1.5,, 200)
# 
# y<- Wendland( dt, k=2, dimension=2)
# 
# plot( dt, y, type="l")
# 
# # should agree with 
# 
# y.test<- Wendland2.2( dt)
# points( dt, y.test)
# 
# # second derivative
# plot( dt, Wendland( dt, k=4, dimension=2, derivative=2), type="l")
# 
# # a radial basis function using the Wendland  the  "knot" is at (.25,.25)
# gl<- list( x= seq( -1,1,,60), y = seq( -1,1,,60) )
# 
# bigD<- rdist( make.surface.grid( gl), matrix( c(.25,.25), nrow=1))
# RBF<- matrix(Wendland( bigD, k=2, dimension=2), 60,60)
# 
# # perspective with some useful settings for shading.
# persp( gl$x, gl$y, RBF, theta =30, phi=20, shade=.3, border=NA, col="grey90")


```


